### 2012-10-29 2.tex
Writing on 2-4 and appendices

\chapter{Gradient and categorical aspects of wordlikeness judgements} 
\label{gradience}

%\citet{Fischer-Jorgensen1952}
%\citet{Saporta1958}
%\citet{Saporta1957}

\citet{Halle1962} and \citet{Chomsky1965} use two nonce words, \emph{blick} [blɪk] and \emph{bnick} [bnɪk], to illustrate that speakers internalize generalizations about possible and impossible words in their language. While \emph{blick} is unattested, but in some sense a ``possible'' word of English, \citeauthor{Chomsky1965} observe that speakers immediately recognize \emph{bnick} to be ill-formed, an impossible word, despite the fact that it is well-formed in other languages (e.g., Moroccan Arabic \emph{bniqa} `closet').

A naïve account of this contrast is based on the assumption that segments or timing units must either be parsed into prosodic structures like the syllable, or be subject to phonological repair. Early precedents for this assumption can be found in work by \citet[10f.]{Hooper1973} and \citet[57f.]{Kahn1976}, and it is fundamental to many later frameworks \citep[e.g.,][]{Ito1989a,Noske1992,OT}.

In English, [bl] is a permissible onset, but unlike some languages (e.g., Moroccan Arabic \emph{bni} `building', \emph{bnat} `daughters', \emph{bniqa} `closet'), [bn] is not. \citet[][19f.]{Wolf2009} suggest, for instance, that an underlying /bnɪk/ would be realized as [nɪk]; it is also possible to imagine resolution by prothesis---[əbnɪk]---or anapytxis---[bənɪk], and indeed \citet{Davidson2006b} finds that English speakers use all three of these repairs when asked to mimic foreign pronunciations of obstruent-nasal onsets like /bn/.\footnote{However, \citet{Davidson2005,Davidson2006a} observes an acoustic and articulatory distinction between ``exscrescent'' echwa found in non-native cluster production and ``lexical'' schwa: the anaptyctic [z${^\textrm{ə}}$k] produced by English speakers attempting to mimic word-initial [zk] is more similar to \emph{scum} than to \emph{succumb}. This may undermine any link between non-native realizations and epenthesis found elsewhere in English (e.g., allomorphs of the regular past tense and noun plural).} However, many linguists now regard the above as overly simplistic, for, as \citeauthor{Shademan2006} argues, such accounts draw a categorical boundary between the possible and impossible which fails to match the granularity at which speakers may report ``wordlikeness'' intuitions.

\begin{quote}
A defect of current grammatical acounts of phonotactics is that they render simple up-or-down decisions concerning well-formedness and cannot account for gradient judgements. But when judgements are elicited in a controlled fashion from speakers, they always emerge as gradient, including all intermediate values. \citep[371]{Shademan2006} 
\end{quote}

This chapter focuses on the hypothesis, implicit in \citeauthor{Shademan2006}'s statement, that there is an inherent gradient or continuum between possible and impossible words, against an null hypothesis that the observed patterns of judgements are little more than an artifact of the wordlikeness task itself. \S\ref{2background} places this debate in a historical perspective, and argues that the naïve but widely accepted form of this hypothesis cannot be falsified. In \S\ref{2evaluation}, a quantitative evaluation of a revised version of this hypothesis reveals a poor match between the gradience in speakers' judgements and contemporary computational models of wordlikeness. Taken together, the findings undermine the gradience hypothesis.

\section{Background}
\label{2background}

\subsection{A brief history of the gradience hypothesis} 
\label{history}

\citeauthor{Chomsky1965} were not the first to consider the notion of possible and impossible words. Their primary contribution is that their mentalist perspective: they recognize that naïve speakers effortlessly acquire language-specific generalizations about possible and impossible words and can report them without any explicit training.

\subsubsection{The Copenhagen school}

Wordlikeness received some attention by structuralists of the Copenhagen school. In an early study, \citet{Fischer-Jorgensen1952} proposes that possible and impossible words represent endpoints on a continuum, and that no non-arbitrary line can be drawn between them. While it would anachronistic to assign this a mentalistic interpretation, this appears to be the earliest proposal that wordlikeness contrasts are in some sense gradient.

However, not all early literature is concerned with gradience. \citet[31]{Vogt1954}, for instance, recognizes that the taxonomic phoneme is insufficient to account for many wordlikeness contrasts. \citeauthor{Vogt1954} observes that allophony may account for the absence of certain phone sequences, but it does not provide a suitable explanation for the absence of initial [bn] in English, nor does it make correct predictions about the surface realization of an underlying initial /bn/. \citeauthor{Vogt1954} concludes that additional grammatical machinery will be needed to account for possible and impossible words.

\subsubsection{Early generative grammar}

Gradient grammaticality was a tenet of the earliest work on generative syntax. For instance, \citet[132]{LSLT} writes that ``there is little doubt that speakers can fairly consistently order new utterances, never previously heard, with respect to their degree of `belongingness' to the language''. \citet{ASPECTS} proposes that different syntactic violations give rise to different degrees of ungrammaticality. Unlike \citeauthor{Fischer-Jorgensen1952}, however, for \citeauthor{LSLT} this continuum does incorporate a hard-and-fast boundary between the grammatical and ungrammatical: there are no constrasts between grammatical utterances, merely between different degrees of ungrammaticality. For \citeauthor{LSLT}, all grammatical utterances are alike with respect to their grammaticality, but there are many degrees of ungrammaticality \citep[61]{Schutze1996}.

\citet{SPR,Halle1962} and \citet{Stanley1967} concern themselves only with categorical wordlikeness contrasts, however. Both authors derive wordlikeness effects from morphophonemic redundancy rules acting on underlying representations. The discussion of \emph{bnick} by \citet[101]{Chomsky1965} is a good example: \citeauthor{Chomsky1965} observe that a consonant before a word-initial stop can only be a liquid in English. A corresponding redundancy rule which specifies a consonant in this position as [$+$\textsc{Liquid}] precludes the derivation of \emph{bnick}, since there is no way to realize the second consonant as a nasal.\footnote{This is not the only redundancy that could rule out \emph{bnick}. For instance, /s/ is the only word-initial consonant that may precede a nasal consonant in English.}

\subsubsection{\emph{The sound pattern of English}} 
\label{2spe}

In \emph{The sound pattern of English} (Henceforth \emph{SPE}), \citet{SPE} argue that a complete solution to wordlikeness must countenance gradience. In addition to the contrast of \emph{blick} and \emph{bnick}, they introduce a third nonce word \emph{bnzk}, which they take to be even less well-formed than ``impossible'' \emph{bnick}. They write:

@@

%In some languages (e.g., Imdlawn Tashlhiyt Berber [tm.zħ] `she jested', [tzmt] `it (f.) is stifling'; \citealt{Dell1985}), \emph{bnzk} is syllabifiable, so its ill-formedness is once again a property of English, not a universal.

\citeauthor{SPE} propose to derive gradience from the complexity (using a simple feature-counting metric) of the redundancy rules needed to map a nonce word to an attested word, and they show that this derives the increasing cline of wordlikeness running from ``consonant soup'' \emph{bnzk} to ill-formed \emph{bnick}, possible \emph{blick}, and finally lexical \emph{brick}.

Despite the considerable attention given to the proposals of \emph{SPE} in the wake of that book's publication in 1968, the \emph{SPE} wordlikeness model has received almost no further attention in the literature. At the risk of explaining what might be no more than an an accidental gap in the literature, the novel aspects of \emph{SPE} model---gradience derived from similarity to existing lexical entries---may have been overshadowed by the many other contentious proposals in \emph{SPE}, and particularly by compelling arguments against the assumption that wordlikeness contrasts derive solely from properties of underlying forms. \citet{Shibatani1973} observes that there are some generalizations about surface forms which give rise to wordlikeness contrasts, but cannot be stated as constraints on underlying forms. An example from German is shown in (\ref{fd}) below.

\begin{example}[German final devoicing] 
\label{fd}
\begin{tabular}{l l l l}
   & nom.sg. & nom.pl.    \\
a. & [piːp]    & [piːpə]  & `cheep(s)'      \\
   & [diːp]    & [diːbə]  & `thief/thieves' \\
b. & [ɡʀaːt]   & [ɡʀaːtə] & `ridge(s)'      \\
   & [ɡʀaːt]   & [ɡʀaːdə] & `degree(s)'     \\
\end{tabular}
\end{example}

\noindent The plural appears to preserve a contrast in final obstruent voicing which is absent in the singular:\footnote{While there is a contentious debate as to whether devoicing in German is completely neutralizing \citep[e.g.,][]{Fourakis1984} or not \citep[e.g.,][]{Port1985}, it is irrelevant to the discussion at hand.} word-final voiced stops are never found in German, and \citet[95]{Shibatani1973} reports that ``it is easy to show that a native speaker of German rejects those forms \emph{on the ground that they end in voiced obstruents}'' (emphasis in original). However, given that root-final voicing is not predictable (e.g., /ɡʀaːt \alt{} ɡʀaːd/), the process of \textsc{Final Devoicing} cannot be any kind of lexical redundancy and is mysterious under the \emph{SPE} account. The alternative proposed by \citeauthor{Shibatani1973} and by \citet{Clayton1976}, is that to say that nonce words like [ɡʀaːd] are ill-formed not because of any underlying property, but since they fail to undergo an otherwise-exceptionless phonological process, \textsc{Final Devoicing}, or equivalently, the surface-true generalization it derives.

\citet{Sommerstein1974} further notes that the computation of hypothetical underlying forms from nonce surface forms which is implied by the \emph{SPE} theory is non-trivial in the presence of phonological opacity \citep[see][528f.]{Anderson1988a}, and the rejection of the biuniqueness principle means that there is not always a unique solution, as is illustrated by the two underlying forms corresponding to [ɡʀaːt] in (\ref{fd}) above.

\subsubsection{Autosegmental phonology and beyond}

The arguments of \citeauthor{Shibatani1973} and others led theorists to focus their attention on properties of surface representations as determinants of wordlikeness. Though syllabification plays no role in \emph{SPE}, it is crucial to many earlier studies \citep[for a review, see][]{Goldsmith2011b}, and it received particular attention in the 1970s. \citet{Hooper1973} and \citet{Kahn1976} argue that the syllable is useful for defining wordlikeness generalizations.\footnote{\citet{Steriade1999} and \citet{Blevins2003}, however, argue that a number of phonotactic generalizations previously stated in syllabic terms can be reanalyzed without making reference to syllables.} \citeauthor{Hooper1973} argues, for instance, that [bn], impossible as an English onset, is unobjectionable as a syllable contact cluster in nonce words like \emph{stabnik} (or in names like \emph{Abner}), and that this demonstrates the superiority of syllable-based wordlikeness generalizations. This already signals further trouble for alternative accounts which focus on underlying forms. Syllabification may span morphs, is generally predictable, and is universally non-contrastive, and as a consequence, few posit in to be present in underlying representations \citep[though see, e.g.,][]{Vaux2003}. \citeauthor{Hooper1973} also points to loanword adaptations which produce native syllable structure \citep[e.g.,][]{Carlisle1991} as evidence that syllabification is part of the phonological computation. Further enrichments to the theory are provided by the autosegmental theory of the syllable \citep{McCarthy1979b}, which envisions the syllable as an articulated tree structure \citep[as first envisioned by][]{Pike1947a}, and theories like prosodic licensing \citep{Ito1989a}, in which syllabification triggers phonological repairs.

While none of these authors discuss gradience, the syllable plays a role in the definition of a gradient measure, ``positional probability'', though to correlate closely with human judgements of wordlikeness. The positional probability of a nonce monosyllabic word is derived from the combined probabilities at which segments occur in onset, nucleus, and coda positions in the lexicon of a given language.

Using the head-term preference paradigm, \citet{Jusczyk1993b} and \citet{Friederici1993} find that typically-developing children as young as 9 months of age distinguish between nonce words which are and are not phonotactically valid in their target language. \citet{Jusczyk1994} report that 9-month-old children acquiring English also show preferences for nonce words with high positional probability over those with low positional probability. Faciliatory effects of positional probability (i.e., shorter latencies) are reported for other nonce word tasks conducted with adults, including single-word shadowing \citep{Vitevitch1997,Vitevitch1998}, same/different judgements \citep{Vitevitch1999a,Luce2001,Lipinski2005,Vitevitch2005}, and lexical decision \citep{Pylkkanen2002a}.

Most relevant to the question at hand, \citet{Frisch2000} and \citet{Vitevitch1997} find that speakers' wordlikeness ratings of multisyllabic words are correlated wtih the positional probailities of the constituent syllables. Unfortunately, none of these researchers make any effort to eliminate the possibility that the low positional probability stimuli are ``impossible'' words of English. In fact, 
Chapter \ref{clusters} argues
%the author has argued elsewhere \citep{Gorman2012c}
that many of the stimuli used by \citeauthor{Frisch2000} and \citeauthor{Vitevitch1997} contain illicit word-medial consonant clusters. While \citeauthor{Vitevitch1997} neither control nor manipulate the well-formedness of medial clusters, in a post-hoc test they consider a probabilistic measure of cluster well-formedness, which reveals that cluster well-formedness is correlated with syllable-internal positional probabilities and wordlikeness judgements, but \citeauthor{Vitevitch1997} ultimately conclude this cannot explain all the variation in wordlikeness.

The aforementioned studies all conclude that the gradient measure of positional probability correlates with behavioral results. As the flaws of the \citet{Vitevitch1997} study demonstrate, the aforementioned studies do little to tease apart the gradient and categorical aspects of phonotactics. More generally, they do little to distinguish between positional probability and closely correlated measures like bigram probability (see \S\ref{bigram} below) or neighborhood density, since these studies carefully select stimuli which either have high or low values for all of positional probability, bigram probablility, and neighborhood density. This is particularly troublesome given that no justification has ever been given for the positional probability measure in the first place; it appears to have been created \emph{ex nihilo}; in contrast, the effects of neighborhood density in various psycholinguistic tasks are emergent properties of many models of speech production \citep[e.g.,][]{Luce1998,Luce2000} and perception \citep{Marslen-Wilson1984,Marslen-Wilson1987,McClelland1986,Norris1994,Norris2000}.

\subsection{A naïve gradience hypothesis}

Recent critiques of this syllabification-based model focus on the existence of intermediate wordlikeness ratings \citep[see also][]{Coleman1997,Anttila2008a}.

\begin{quote}
In the particular domain of phonotactics gradient intuitions are pervasive: they have been found in every experiment that allowed participants to rate forms on a scale.
\citep[][382]{Hayes2008a}

When native speakers are asked to judge made-up (nonce) words, their intuitions are rarely all-or-nothing. In the usual case, novel items fall along a gradient cline of acceptability. \citep[][9]{Albright2009a}
\end{quote}

As discussed in \S\ref{2spe} above, this is also the view of \citeauthor{SPE} in \emph{SPE}. The observation itself cannot be questioned, but \citeauthor{SPE}, \citeauthor{Hayes2008a}, and \citeauthor{Albright2009a} do not explicitly state why this data is relevant to the construction of models of wordlikeness. Presumably, these authors believe that these patterns of judgements demonstrate that wordlikeness, as an internal state, is gradient simply because subjects make use of intermediate degrees of wordlikeness in judgement tasks. This proposition, generalized below, is ``naïve'' not because it lacks sophistication, but because it is rooted in a belief in naïve realism, a philosophy which holds that perception provides a relatively direct picture of the nature of the world, an influential view in the cognitive sciences in general (see \citealt{Fodor1981a} for a critique).

\begin{unlabeledexample}
\textsc{Naïve Gradience Hypothesis}: If graded judgements of a concept or mental state use intermediate ratings, that concept or mental state is gradient
\end{unlabeledexample}

This proposition has the form of an indicative conditional. While the antecedent of this condition, the presence or absence of intermediate ratings in a graded judgement task, is readily observable, it has no bearing on the truth or falsity of the conditional proposition. However, whereas a false consequent (e.g., an all-or-nothing concept) would potentially falsify the proposition, the consequent, a mental state, cannot be directly observed. If this is the case, then both the proposition and the consequent (the proposition that wordlikeness is itself gradient) fail the test of falsification, placing it beyond the scope of scientific inquiry.

\citet{Armstrong1983} cut the Gordian knot by asserting the possibility of observing a false consequent; they find the existence of all-or-nothing concepts like ``odd number'' or (more controversially) ``female'' to be apparent.

\begin{quote}
Are there definitional concepts? Of course. For example, consider the concept \emph{odd number}. This seems to have a clear definition, a precise description. [\ldots{}] No integer seems to sit on the fence, undecided as to whether it is quite even, or perhaps a bit odd. No odd number seems odder than any other odd number. \citep[274]{Armstrong1983}
\end{quote}

\citeauthor{Armstrong1983} note that if subjects produce intermediate ratings for these categories, then the naïve proposition is false, and this is what \citeauthor{Armstrong1983} find in their experiments. They ask subjects to rate, on a seven-point scale, the extent to which, e.g., certain odd counting numbers represent the concept ``odd number'', and find that speakers do make use of intermediate values, as shown in Figure \ref{agg}. In summary, the proposition that intermediate ratings indicate mental gradience is either untestable, or it is false, and is thus rejected.

\begin{figure} 
\centering
\includegraphics{agg.pdf}
\caption{Subjects asked to rate the degree to which certain even and odd numbers represent the categories ``even'' and ``odd'', respectively, freely use intermediate ratings.}
\label{agg} 
\end{figure}

\subsection{A falsifiable gradience hypothesis}

\citeauthor{Armstrong1983} do not view their results as evidence that subjects have an internal model of odd numbers which is at odds with the formal, all-or-nothing definition, and in fact, several additional experiments show that subjects deploy the extension of the formal definition in other tasks. Reviewing this study, \citet[215]{Schutze2011} writes that the experiments show that judgements are ``sensitive to factors other than our underlying competence''. In the case of wordlikeness, a case can be made that one such factor is similarity to existing words (e.g., \citealt{Bailey2001}, \citealt[][151, fn. 27]{LSLT}, \citealt{Greenberg1964}, \citealt{Ohala1986b}, \citealt{Schutze2005}, \citealt{Sendlmeier1987}, \citealt{Vitz1973}). However, the mere fact that subjects use intermediate ratings does not show that they do so with any systematicity, or that, e.g., the contrast between \emph{447} and \emph{7} with respect could be given any satisfying explanation. \citeauthor{Schutze2011} further suggests that gradient responses are the product of the task itself:

\begin{quote}
Putting it another way, when asked for gradient responses, participants will find some way to oblige the experimenter; if doing so is incompatible with the experimenter's actual question, they apparently infer that she must have really intended to ask something slightly different. \citep[215]{Schutze2011}
\end{quote}

This introduces a troubling possibility, that subjects ``oblige'' experimenters by introducing random noise to their categorical judgements when presented with a Likert task. Were this the case, the observed intermediate judgements are unlikely to yield information about speakers' knowledge of possible and impossible words. The alternative considered here is that there exists some model in which the gradience in ratings is systematic.

\begin{unlabeledexample}
\textsc{Gradience Hypothesis}: If graded judgements of a concept or mental state use intermediate ratings, there is some model which can predict this behavior
\end{unlabeledexample}

\noindent There are now many computational models of wordlikeness which proport to do just this. An unfortunate defect of prior evaluations of these models, however, is that they adopt the naïve gradience hypothesis. \citet[382]{Hayes2008a} state that consider the ability to model gradient intuitions to be so important that they do not include any categorical models in their evaluation, as do many other authors. As a result, the literature contains no serious attempt to evaluate the gradience hypothesis.

@@

The remainder of this chapter is devoted to testing the revised hypothesis. This is accomplished by an evaluation of computational models of wordlikeness in which categorical and gradient models are placed on an equal footing.

\subsection{Data sources}

The evaluation makes exclusive use of previously published English wordlikeness data. For inclusion in the evaluation, a study must conform to all three of the following conditions. First, the stimuli must consist only of monosyllabic words presented auditorily. Secondly, a significant portion of the stimuli must contain gross phonotactic violations (e.g., \emph{bnick}). Finally, the ratings, averaged across subjects, must be publicly available. The large-scale studies by \citet{Bailey2001} and \citet{Shademan2006,Shademan2007} are ineligible because the former lack stimuli with gross phonotactic violations and the latter data are not available to the public in any form. All data are reproduced in Appendix \ref{appendixA}.

\subsubsection{\citealt{Greenberg1964}}

\citet{Greenberg1964} investigated wordlikeness using the technique of free magnitude estimation, a mechanism which has become increasingly popular among syntacticians \citep[e.g.,][]{Bard1996}. At the beginning of the experiment, the subject heard a recording of the word \emph{stick}. In subsequent trials, the subjects heard a nonce word and were asked to report ``how far would you say that is from English?'', with \emph{stick} at ``1''; subjects are told that a word that is ``twice as far from English'' as \emph{stick} should be scored ``2''. The data used here are from \citeauthor{Greenberg1964}'s Experiment B, in which 17 undergraduates were presented 17 stimuli in all. In addition to \emph{stick}, the stimuli include three other English words; these four items were excluded from further analyses, leaving 13 stimuli. As is standard practice in psychophysics \citep[e.g.,][]{Butler1987}, magnitudes were log-transformed prior to analysis.

\subsubsection{\citealt{Scholes1966}}

\citet{Scholes1966} conducted a number of English wordlikeness judgement with middle school children. The data used here come from his ``experiment 5'', in which 63 monosyllabic items were presented to 33 seventh-grade students. For each stimulus, the subjects produced forced choice ``yes''/``no'' answers to the question of whether the item ``is likely to be usable as a word of English''. \citet{Hayes2008a} and \citet{Albright2009a} analyze this data as gradient by performing an item averaging using the fraction of ``yes'' answers for each stimulus \citep[see also][]{Pierrehumbert1994,Coleman1997,Frisch2000}. For instance, 22 of the 33 students answered ``yes'' for \emph{shlerk} [ʃlɚk], so it is assigned a score of $0.666$.\footnote{This procedure conflates intraspeaker variation, which in some cases is considerable \citep{Shademan2007} with speaker-internal gradience, and its adoption here should not be construed as an endorsement. Rather, it is provided solely for comparison with prior studies using the \citeauthor{Scholes1966} data.} These stimuli are all ostensibly non-words, but include \emph{clung} [klʌŋ], the preterite and past participle of the verb \emph{cling}, and \emph{brung} [brʌŋ], a dialectical past participle for \emph{bring}. These two words were excluded and the remaining 61 stimuli were submitted to analysis.

\subsubsection{\citealt{Albright2003b}}

\citet{Albright2003b} gathered wordlikeness judgements to serve as norms for a \emph{wug}-test. 87 items were presented to 24 undergraduate subjects, who rated each word on a seven-point Likert scale. The lowest point on the scale was labeled ``completely bizarre, impossible as an English word'', and that the highest point was labeled ``complete normal, would make a fine English word''.

\begin{table}
\label{counts}
\centering
\begin{tabular}{l r r r}
\toprule
        & subjects & items & observations \\
\midrule
Scholes & 33       & 66    & 2,079        \\
A\&H    & 24       & 88    & 2,112        \\
\bottomrule
\end{tabular}
\caption{Idk.}
\end{table}

\subsection{Model comparison}

The four models used here consist of a binary baseline and three computationally implemented gradient models. These three models are chosen because prior studies have shown they are correlated with wordlikeness judgements. Non-parametric rank correlation statistics are used evaluate the correlation between model predictions and wordlikeness ratings. Rank correlation are used rather than the parametric Pearson correlation statistic that is used by \citet{Hayes2008a}, for instance, because they make none of the potentially troublesome assumptions of the latter method \citep[see][23, fn. 12]{Albright2009a}.

The Spearman $\rho$ is most widely known rank correlation statistic, but it is difficult to give a natural interpretation to this quantity. On the other hand, the Kendall $\tau_b$ and Goodman-Kruskal $\gamma$ can be interpreted as fractions of the number of \emph{concordant} and \emph{discordant} pairs \citep{Noether1981}. Consider the case here, in which model scores are compared with wordlikeness ratings. If a model rates \emph{dresp} [dɹɛsp] more wordlike than *\emph{srest} [sɹɛst] and if speakers further rate \emph{dresp} more English-like than \emph{srest}, than \emph{dresp}/\emph{srest} are a concordant pair. If, however, a model and speakers disagree on the relative ranking of \emph{dresp} and \emph{srest}, the pair is discordant. The $\tau_b$ and $\gamma$ differ only in their treatment of ``ties'' (i.e., if e.g., *\emph{dresp} and *\emph{srest} are scored the same, or rated the same); $\tau_b$ includes a correction for ties, whereas $\gamma$ ignores tied pairs. Much like the familiar Pearson correlation, $\rho$, $\tau_b$ and $\gamma$ are all in the range [$-1$, $1$]. Correlations of the four models are given in Table \ref{cor}.

\subsubsection{Binary baseline}

The binary baseline used here is a crude implementation of the null hypothesis that there are no gradient effects in wordlikeness judgements. To create such a baseline, it is necessary to distinguish between nonce words which contain a gross phonotactic violation and those which do not. As all stimuli here are monosyllables, this task can be further simplified by separately considering the two major subcomponents of the syllable, the onset and rime. Speakers are particularly adept as separating onsets from rimes \citep{Treiman1986,Treiman1995,Fowler1993}, and a large portion of phonotactic generalizations and violations can be localized to either onset or rime \citep[e.g.,][]{Borowsky1989,Fudge1969,Treiman1995,Kessler1997,Treiman2000}.

The baseline considers a nonce syllable to be well-formed if it consists of both an attested onset and an attested rime; the free combination of these two components is the only mechanism by which this model can generalize beyond attested words. It is surely the case that this is an ultimately insufficient model: \citet{Albright2009a} observes, for instance, that [ɛsp] is a well-formed rime, even though it is found in no word of English. This problem is more acute in languages with more permissive syllable structures than those of English. For instance, \citet{Fischer-Jorgensen1952} and \citet{Vogt1954} assert that there are many accidental gaps (i.e., possible but unattested structures) in the inventory of consonant clusters in Georgian, a language which admits as many as five adjacent consonants.\footnote{
Chapter \ref{clusters} 
%\citet{Gorman2012c}
reports a similar result for English syllable contact clusters.} Despite these defects, the baseline outperforms the gradient models in most contexts.

The rating densities from the three studies, linearly transformed to fit within the interval [0, 1] and split according to this binary baseline, are shown in Figure \ref{dsn}. In all three studies, it is possible to discern a relatively sharp separation between valid and invalid clusters, but also the presence of intermediate values.

\begin{figure} 
\centering
\includegraphics{density.pdf}
\caption{Average ratings of individual nonce words, linearly transformed to fit the interval [0, 1], tend to clump into two groups with little overlap; words which consist of  attested onsets and rimes receive ratings near ceiling, whereas ratings of phonotactically invalid words are spread across the lower half of the spectrum.}
\label{dsn}
\end{figure}

\begin{table} 
\centering
\begin{tabular}{l l l l l}
\toprule
Spearman $\rho$          & baseline         & maxent  & bigram           & density          \\
\midrule
\citealt{Greenberg1964}  & {0.845}          & {0.765} & {\textbf{0.863}} & {0.648}          \\
\citealt{Scholes1966}    & {0.791}          & {0.762} & {\textbf{0.827}} & {\textbf{0.827}} \\
\citealt{Albright2003b}  & {0.725}          & {0.429} & {0.708}          & {\textbf{0.742}} \\
\midrule
Kendall $\tau_b$         & baseline         & maxent  & bigram           & density          \\
\midrule
\citealt{Greenberg1964}  & {\textbf{0.716}} & {0.585} & {0.692}          & {0.462}          \\
\citealt{Scholes1966}    & {\textbf{0.664}} & {0.597} & {0.652}          & {0.565}          \\
\citealt{Albright2003b}  & {\textbf{0.599}} & {0.343} & {0.506}          & {0.556}          \\
\midrule
Goodman-Kruskal $\gamma$ & baseline         & maxent  & bigram           & density          \\
\midrule
\citealt{Greenberg1964}  & {\textbf{1.000}} & {0.684} & {0.692}          & {0.462}          \\
\citealt{Scholes1966}    & {\textbf{0.995}} & {0.634} & {0.667}          & {0.614}          \\
\citealt{Albright2003b}  & {\textbf{0.953}} & {0.656} & {0.509}          & {0.575}          \\
\bottomrule
\end{tabular}
\caption{Rank correlations between wordlikeness ratings and phonotactic models surprisingly reveal that the binary baseline meeds or exceeds the coverage of three state-of-the-art phonotactic models. All correlations are significant at $p = 0.05$.}
\label{cor}
\end{table}

\subsubsection{Maximum entropy phonotactics}

\citeauthor{Hayes2008a} (\citeyear{Hayes2008a}; henceforth H\&W) develop a sophisticated model of phonotactic grammaticality which estimates a probability distribution over phoneme sequences by weighing constraints according to the principle of maximum entropy, following \citet{Goldwater2003} and \citet{Jager2007}. H\&W report that the predictions of their model are closely correlated with the \citet{Scholes1966} wordlikeness ratings. A direct replication of their predictions was attempted by using the software, model parameters, and training data as described in that study. Since the training of the maximum entropy model is inherently stochastic, producing slightly different outcomes on each run, the lowest scoring of ten runs is reported (H\&W:396), though in general there is not a great deal of variation between individual runs. One limitation of this model is that it is not feasible to score whole words, as the number of constraints which must be inspected grows exponentially as the span of possible constraints increases. Following H\&W and of \citet{Albright2009a}, who also applies the maximum entropy model to the \citet{Albright2003b} norms, the model is trained and scored only on stimulus onsets. However, as a consequence, the maximum entropy model performs particularly poorly on this data set, as many stimuli contain phonotactic violations in rime positions.

\subsubsection{Segment bigram probability} 
\label{bigram}

The bigram probability of a sequence $ijk$ is the product of the probability of an sequence-initial $i$, the probability that $j$ follows $i$, and the probability that $k$ follows $j$, and the product of sequence-final \emph{k}.

\begin{unlabeledexample}
$\displaystyle \hat{p}(ijk) = p(i|\textrm{start}) \cdot p(j|i) \cdot p(k|j) \cdot p(\textrm{stop}|k)$
\end{unlabeledexample}

\noindent \citet{Albright2009a} employs bigram probability to model wordlikeness judgements. While the focus of \citeauthor{Albright2009a}'s study is on developing a model which uses bigrams over phonological features rather than segments themselves, \citeauthor{Albright2009a}'s evaluation, which includes both the \citeauthor{Scholes1966} and \citeauthor{Albright2003b} data sets, finds an advantage for segmental bigrams. \citeauthor{Albright2009a} does not provide an implementation of the featural bigram model, nor does his study describe it in sufficient detail to allow for a new implementation, but segmental bigram scores for the \citeauthor{Albright2003a} data are reported in the appendix. As reported by \citeauthor{Albright2009a}, segmental bigrams outperform featural bigrams (see Table \ref{albrightimproved}).

>>>>

\chapter{Categorical and gradient aspects of wordlikeness judgements} 
\label{gradience}

Two nonce words, \emph{blick} [blɪk] and \emph{bnick} [bnɪk], underlie the claim that speakers can rapidly distinguish between possible and impossible words \citet{Halle1962}. In \emph{SPE}, \citet{SPE} use a third word, \emph{bnzk}, to argue that nonce words fall on a cline of well-formedness. Neither \emph{bnick} nor \emph{bnzk} are possible words of English, they are possible words in other languages: stop-nasal onsets are found in Moroccan Arabic (e.g., \emph{bniqa} `closet') and stop-nasal-fricative-stop words in Imdlawn Tashlhiyt Berber \citep{Dell1985}. However, there is some sense in which \emph{bnzk} is even less English-like than \emph{bnick}. Of this, \citeauthor{SPE} write:

@@

\noindent
The position that the well-formedness of nonce words is consistent the view of syntactic grammaticality taken in \emph{LSLT} and \emph{Aspects} \citep{LSLT,ASPECTS}, where it is claimed that different syntactic vioaltions result in different degrees of ungrammaticality.

Most recent discussions of the ``problem of admissibility'' focus on this cline of wellformedness as it is evidenced in wordlikeness tasks.

\begin{quote}
When native speakers are asked to judge made-up (nonce) words, their intuitions are rarely all-or-nothing. In the usual case, novel items fall along a gradient cline of acceptability. \citep[][9]{Albright2009a}

In the particular domain of phonotactics gradient intuitions are pervasive: they have been found in every experiment that allowed participants to rate forms on a scale.
\citep[][382]{Hayes2008a}

\ldots{}when judgements are elicited in a controlled fashion from speakers, they always emerge as gradient, including all intermediate values. \citep[371]{Shademan2006} 
\end{quote}

At first blush, this would seem to argue against a naïve account of wellformedness judgements in which ill-formedness results when prosodic parsing fails (e.g., \citealt{Ito1989a}, \citealt{Noske1992}, \citealt{OT}). Consider the possibility that a nonce word is ill-formed if it cannot be syllabified without modification. There is a long history for the proposal that impossible words are those whose surface form cannot be parsed into prosodic structures like syllables without further phonological modification (e.g., \citealt[10f.]{Hooper1973}, \citealt[57f.]{Kahn1976}). There is no doubt that prosodic context is involved in determining where segments may occur. For instance, [ŋ] is a possible coda in English, but not a possible onset. Similarly, the [bn] of \emph{bnick} is judged to be admissible (though unattested) in medial position in English (e.g., \emph{sta}[b.n]\emph{ick}; \citealt[97]{Hooper1973}) but not as an onset. English is unable to sylalbify initial onset /bn/ without further modification, so one might expect underlying /bnɪk/ to be subject to deletion \citep[19f.]{Wolf2009}, prothesis, or anaptyxis. Indeed, \citet{Davidson2006b} finds that English speakers use all three of these repairs when asked to mimic foreign pronunciations of obstruent-nasal onsets impossible in English.\footnote{However, \citet{Davidson2005,Davidson2006a} finds acoustic and articulatory distinctions between the ``exscrescent'' schwa used to resolve non-native onset clusters, and the ``lexical'' schwa found in real words. For example, the anaptyctic [z${^\textrm{ə}}$k] produced by English speakers attempting to mimic word-initial [zk] more closely resembles the onset of \emph{scum} than the start of \emph{succumb}. This may undermine any link between non-native realizations and epenthesis of schwa in native words (e.g., in the syllabic allomorphs of the regular past tense and noun plural).}

However, prosodic well-formedness distinctions do not provide a ready explanation for the ``numerous other distinctions'' posited in \emph{SPE}. This is thought to rule out simple prosodic account of wordlikeness judgements.

\begin{quote}
A defect of current grammatical acounts of phonotactics is that they render simple up-or-down decisions concerning well-formedness and cannot account for gradient judgements. \citep[371]{Shademan2006}
\end{quote}

Implicit in this claim is the assumption that the graded judgements obtained from wordlikeness tasks which permit gradient responses require a gradient model of well-formedness. However, it is argued here this is not pertinent to the question of whether wordlikeness is gradient or categorical, because graded judgements are a property of all graded tasks, even when applied to judgements of discrete categories.

\citet[382]{Hayes2008a} ``consider the ability to model gradient intuitions to be an important criterion for evaluation phonotactic models''. However, they fail to ask consider whether categorical models provide a reasonable fit to gradient judgements. Rather than directly evaluating the categorical model of English phonotactics proposed by \citet{Clements1983}, \citeauthor{Hayes2008a} convert these constraints, many of which are exceptionless, into probabilitistic constraints. This suggests that \citeauthor{Hayes2008a} equate ``the ability to model gradient intuitions'' with the ability to make continuous predictions about well-formedness. But if gradient judgements are nothing more than a task effect, a model need not make gradient predictions to fit gradient judgements.

Despite many claims that categorical models are empirically insufficient, there is not a single quantitative investigation of categorical wordlikeness models, and no study which compares categorical and gradient wordlikeness models on an equal footing. This chapter represents a first attempt to fill that lacuna.

OUTLINE HERE

\section{What wordlikeness might not be}

\subsection{Gradience in wordlikeness judgements}

\subsection{Gradience as a task effect}

@@

\subsection{Method}

\subsubsection{Materials and subjects}

\subsubsection{Computational models}

\subsection{Results}

\subsection{Discussion}
